---
title: " World Development dataset analysis"
author: "Peter Guo"
date: "2025-2-26"
format:
  html:
    code-fold: true
  pdf: default
bibliography: references.bib
---

## 2. Load the dataset using R.
```{r}
library(tidyverse)
wdi_data <- read.csv("wdi.csv", header = TRUE)
```

## 3.Conduct exploratory data analysis on at least three indicators of your choice. Summarise your findings in markdown sections.
- Summary Statistics
```{r}
# Summary for GDP per Capita
gdp_per_capita_summary <- summary(wdi_data$gdp_per_capita)
gdp_per_capita_summary

# Summary for Life Expectancy
life_expectancy_summary <- summary(wdi_data$life_expectancy)
life_expectancy_summary

# Summary for Total Population
total_population_summary <- summary(wdi_data$total_population)
total_population_summary
```
The summary provides key metrics (minimum, 1st quartile, median, mean, 3rd quartile, maximum) for each indicator. This helps us understand the central tendency and spread of the data.

- Correlation Analysis
```{r}
# Calculate correlation matrix
cor_matrix <- wdi_data %>% 
  select(gdp_per_capita, life_expectancy, total_population) %>% 
  cor(use = "complete.obs")
print(cor_matrix)
```
The correlation matrix shows the linear relationships between the variables. For example, a positive correlation between GDP per Capita and Life Expectancy might suggest that countries with higher incomes tend to have longer life expectancies.

The correlation matrix shows a moderate positive relationship between GDP per capita and life expectancy (r ≈ 0.64), indicating that higher income levels are generally associated with longer life expectancies. In contrast, both GDP per capita and life expectancy have almost no relationship with total population (r ≈ -0.06 and r ≈ -0.01, respectively), suggesting that a country's population size does not significantly impact these measures.

## 4.Create at least two different types of plots (e.g., bar chart, scatter plot) to represent your analysis. Use Quarto code chunks to embed these visualisations. Add a title and axis labels to each plot. Use Quarto to include a caption and a reference to the source of the data. Hide your code in the final document.

- top 20 countries with the highest GDP per Capita
```{r, echo=FALSE, fig.cap="Figure 1: Bar chart of GDP per Capita by Country (Top 20, 2022, in ascending order). Data source: World Development Indicators.", `fig-label`="gdp_bar"}
library(ggplot2)
library(dplyr)

# Select top 20 countries with the highest GDP per Capita and arrange them in ascending order
top20 <- wdi_data %>% 
  arrange(desc(gdp_per_capita)) %>% 
  head(20) %>% 
  arrange(gdp_per_capita)

# Create the bar chart
gdp_bar <- ggplot(top20, aes(x = reorder(country, gdp_per_capita), y = gdp_per_capita)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "GDP per Capita by Country (Top 20, 2022)",
       x = "Country",
       y = "GDP per Capita (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

gdp_bar

# save the plot for external use
ggsave("gdp_bar.png", gdp_bar, width = 10, height = 6)


```
- Scatter plot of GDP per Capita vs. Life Expectancy
```{r, echo=FALSE, fig.cap="Figure 2: Scatter plot of GDP per Capita vs. Life Expectancy (2022). Data source: World Development Indicators.", `fig-label`="scatter_plot"}

scatter_plot <- ggplot(wdi_data, aes(x = life_expectancy, y = gdp_per_capita)) +
  geom_point(color = "darkred") +
  labs(
    title = "GDP per Capita vs Life Expectancy (2022)",
    x = "Life Expectancy (Years)",
    y = "GDP per Capita (USD)"
  ) +
  theme_minimal()

scatter_plot


ggsave("scatter_plot_switched.png", scatter_plot, width = 10, height = 6)

```

## 5. Construct a table that highlights some key statistics from your analysis. Ensure the table is well-formatted and included in the report.
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="data source:World Development Indicators.", `fig-label`="stats_table"}
library(dplyr)
library(gt)

stats_gdp <- wdi_data %>%
  summarise(
    Median = median(gdp_per_capita, na.rm = TRUE),
    Variance = var(gdp_per_capita, na.rm = TRUE),
    Mean = mean(gdp_per_capita, na.rm = TRUE)
  ) %>%
  mutate(Indicator = "GDP per Capita")

stats_life <- wdi_data %>%
  summarise(
    Median = median(life_expectancy, na.rm = TRUE),
    Variance = var(life_expectancy, na.rm = TRUE),
    Mean = mean(life_expectancy, na.rm = TRUE)
  ) %>%
  mutate(Indicator = "Life Expectancy")


stats_combined <- bind_rows(stats_gdp, stats_life) %>%
  select(Indicator, Median, Variance, Mean)


stats_combined %>%
  gt() %>%
  fmt_number(
    columns = c(Median, Variance, Mean),
    decimals = 0
  ) %>%
  tab_header(
    title = "Key Statistics for Selected Indicators (2022)",
    subtitle = "Data source: World Development Indicators"
  )


```

## 6. Include cross-references to your figures and tables within the text. Demonstrate proper labeling and referencing techniques.

In this report, we presented a bar chart (Figure @ref(gdp_bar)) that displays the top 20 countries by GDP per Capita, a scatter plot (Figure @ref(scatter_plot)) that illustrates the relationship between Life Expectancy and GDP per Capita, and a summary table (Table @ref(stats_table)) showing the key statistics for these indicators. The cross-references ensure that readers can easily navigate to the relevant figures and table for further details.

## 7. Add a bibliography using BibTeX (.bib). Cite at least two sources related to your analysis.
This report cites the findings from [@Smith2021] and [@Doe2020].
